---
title: "Political and Environmentalist Self-identification"
author: "Katie Munster"
date: "11/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(broom)
library(kableExtra)
```

### Introduction

This report provides analyzes survey data gathered by Chelsea Batavia and Andrews Forest LTER to explore "Influences on charitable giving for conservation." We will explore the following question: Are political affiliation and identifying as an environmentalist independent? This report uses a chi-square test to determine if there is a significant association between political affiliation and identifying as an environmentalist.

```{r}
# Read in data:
conservation_survey_results <- read_csv(here("data", "conservation_survey.csv"))
```

```{r}
conservation_simplified <- conservation_survey_results %>% 
  clean_names() %>% 
  select(env_id, pol_or) %>% # Only keep ENV_ID and POL_OR columns
  mutate(political_affiliation = case_when( # Add a new column for political affiliations
    pol_or == 1 ~ "Conservative",
    pol_or == 2 ~ "Conservative",
    pol_or == 3 ~ "Conservative",
    pol_or == 5 ~ "Liberal",
    pol_or == 6 ~ "Liberal",
    pol_or == 7 ~ "Liberal"
  )) %>% 
  filter(political_affiliation %in% c("Conservative", "Liberal")) # Exclude other political affiliations
```

```{r}
# Counts of liberal and conservative survey respondents who responded "Yes", "No", and "Unsure" to the statement "I consider myself an environmentalist."
environmentalist_counts <- conservation_simplified %>% 
  janitor::tabyl(political_affiliation, env_id)

# Table of both counts and proportions
environmentalist_proportions <- environmentalist_counts %>% 
  adorn_percentages() %>% 
  janitor::adorn_pct_formatting(digits = 2) %>% 
  adorn_ns() 

environmentalist_proportions_kable <- environmentalist_proportions %>% 
  kable(col.names = c("Political affiliation",
                      "Yes",
                      "No",
                      "Unsure"),
        caption = "Counts and proportions of conservative and liberal survey respondants who responded to the statement 'I consider myself an environmentalist.'") %>% 
  kable_styling()

environmentalist_proportions_kable # Print the table
```

```{r}
# Make contingency table. Convert entries in a column to stored rownames
environmentalist_ct <- environmentalist_counts %>% 
  column_to_rownames(var = "political_affiliation")
```

```{r}
# Perform a chi-square test for independence
environmentalist_x2 <- chisq.test(environmentalist_ct)
```

```{r}
environmentalist_tidy <- tidy(environmentalist_x2)
```

There is a significant association between political association (conservative or liberal) and identifying as an environmentalist ($\chi$^2^(`r environmentalist_tidy$parameter`) = `r round(environmentalist_tidy$statistic,2)`, *p* = `r round(environmentalist_tidy$p.value, 3)`). The majority of conservatives (`r environmentalist_proportions$"2"[1]`) did not consider themselves an environmentalist, and the majority of liberals (`r environmentalist_proportions$"1"[2]`) did identify as an environmentalist.


### Data Citation

Andrews Forest LTER Site and C. Batavia. 2019. Influences on charitable giving for conservation: Online survey data of 1,331 respondents across the US, August 2017 ver 3. Environmental Data Initiative. https://doi.org/10.6073/pasta/926e6270e324a1322a900da14d38b96c.

